Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2022-06-07 inst/manuscript/manuscript_microxanox.tex
Fix crossrefs

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header
%% \documentclass[preprint,12pt, a4paper]{elsarticle}

@@

\begin{document}
\begin{frontmatter}

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug\corref{Corresponding
Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey}
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \address[University of Zürich]{Department of Evolutionary Biology
and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  Ca. 100 words.
  \end{abstract}
   \begin{keyword} reproducibility; regime shift;\end{keyword}
 \end{frontmatter}

@@

\begin{itemize}
\tightlist
\item
  \textbf{\textcolor{red}{update links}}
\item
  \textbf{\textcolor{red}{update cross references}}
\item
  \textbf{\textcolor{red}{check keywords}}
\item
  \textbf{\textcolor{red}{add more references}}
\item
  \textbf{\textcolor{red}{expand on Impact and Conclusion section}}
  \pagebreak
\end{itemize}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.05}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.41}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.54}}@{}}
\toprule
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Please fill
in this column} \\
\midrule
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & CC BY 4.0 \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
& & \\
& & \\
& & \\
& & \\
& & \\
C8 & If available Link to developer documentation/manual &
\textcolor{red}{TO BE ADDED} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Rainer@krugs.de \\
\bottomrule
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
(\protect\hyperlink{ref-Binzer2016a}{Binzer et al., 2016};
\protect\hyperlink{ref-Chaparro-Pedraza2021}{Chaparro-Pedraza, 2021};
\protect\hyperlink{ref-Vasseur2005}{Vasseur and McCann, 2005}). They are
also important for developing hypotheses that can then be tested in
empirical studies. One area of ecology in which simple models have
played a influential role is the area of how ecosystems respond to
gradual change in an environmental driver
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). An
environmental driver is a environmental condition that affects an
ecosystem, but is assumed to not be affected by the ecosystem, such as
the rate of nutrient input into a lake.

It is conceivable that an ecosystem state, such as the total biomass of
a particular type of bacteria, may remain unchanged when an
environmental driver changes. It is also possible that the ecosystem
state changes gradually. It is also possible that the ecosystem state
changes abruptly to a new state that is difficult to recover from
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). This
possibility for abrupt, perhaps catastrophic changes that are difficult
to reverse causes considerable concern
(\protect\hyperlink{ref-Collins2021}{Collins et al., 2021};
\protect\hyperlink{ref-Northrop2021}{Northrop et al., 2021};
\protect\hyperlink{ref-Vandermeer2019}{Vandermeer and Perfecto, 2019}).

An example where a gradual change of an environmental variable causes an
abrupt change of the system is the switch from an aerobic (oxygen is
available for metabolism) to anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. Three types of microbes occur in the
model: cyanobacteria (CB) dominating the oxic state, and two types of
sulfur bacteria that dominate the anoxic state (sulfate reducing
bacteria (SB) and phototrophic sulfur bacteria (PB)). This system has
been investigated by \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) in a simulation study of a
mathematical model. They showed that gradual change in the rate at which
oxygen could diffuse into the ecosystem (termed the oxygen
diffusitivity) could cause catastrophic changes in the ecosystem state
that would be difficult to reverse.

One feature of the study by \protect\hyperlink{ref-Bush2017}{Bush et
al.} (\protect\hyperlink{ref-Bush2017}{2017}) was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves open the question of if and how biodiversity
within these types (i.e.~functional groups) of bacteria, in their model,
affects the ecosystem response to environmental change. This limitation
is not specific to the study of \protect\hyperlink{ref-Bush2017}{Bush et
al.} (\protect\hyperlink{ref-Bush2017}{2017}), there are few if any
studies of the effects of biodiversity on abrupt transitions between
ecosystem states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change \protect\hyperlink{ref-Limberger2022}{Limberger}
(\protect\hyperlink{ref-Limberger2022}{2022}), and to base our work on
the work and model of \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}). It was with this goal in mind
that we developed the \emph{microxanox} package. The first stage of
development was to write code from scratch (as there was no available
code to start from) and to confirm that this new implementation would
reproduce the previously published results. The resulting reproduction
is available as one of the package vignettes:
\href{LINK\%20NEEDED}{vignette Partial reproduction of Bush et al}.

@@

\begin{quote}
\{\textgreater\textgreater Describe the software in as much as is
necessary to establish a vocabulary needed to explain its
impact.\textless\textless\}
\end{quote}

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on
\protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) (See
\protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) for a detailed discussion of
the model). At the core of the simulations is a set of ordinary
differential equations (specified in the function
\texttt{bushplus\_dynamic\_model()}, though this function need not be
directly called). There are functions for running individual simulations
and for running a set of simulations across, for example, a range of
environmental conditions.

@@

\begin{quote}
\{\textgreater\textgreater Give a short overview of the overall software
architecture; provide a pictorial component overview or similar (if
possible). If necessary provide implementation
details.\textless\textless\}
\end{quote}

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it easy to reproduce results (see
the supplement to \protect\hyperlink{ref-Limberger2022}{Limberger}
(\protect\hyperlink{ref-Limberger2022}{2022}) for an example of how this
is used). As such, all the parameters needed to run a simulation or find
a stable state (i.e.~the final state of the ecosystem) are contained in
a single object (which can easily be created using included functions).
This parameter object is given to a function that runs the simulations
and returns the results. The returned results object is identical to the
parameter object but with an additional slot named \texttt{results},
which contains the results of the run. Thus the returned results object
contains the simulation conditions (parameters) as well as the results,
and can be used to run the simulation again. This promotes
reproducibility and makes incremental changes of individual parameters
and re-running the simulations straightforward.

@@

A typical simulation would look as shown in @ref(fig:runsim\_example).

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package
(\protect\hyperlink{ref-Soetaert2010}{Soetaert et al., 2010}). The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains
among other things the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed please see the
\emph{User Guide} which accompanies the package or is available at
\href{@LINK_NEEDED}{User Guide}

@@

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) study. The second runs two
simulations, with step-wise and slowly temporally increasing oxygen
diffusivity, and the other with step-wise and slowly decreasing or
decreasing oxygen diffusivity. During this temporal environmental
change, the state of the system was recorded just before change to a new
oxygen diffusivity (we term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

The replication method is implemented in the function
\texttt{run\_replication\_ssfind()} which takes a parameter object as
returned by the function \texttt{new\_replication\_ssfind\_parameter()}
and the number of cores for multithreading the simulation. As the
multithreading uses the package function \texttt{mclapply()} from the
package \texttt{parallel} (\protect\hyperlink{ref-RCoreTeam2022}{R Core
Team, 2022}), the multthreading only works on Linux and Mac. It is
planned to move to \texttt{parLapply()}
(\protect\hyperlink{ref-RCoreTeam2022}{R Core Team, 2022}) in a future
release.

@@

The function \texttt{plot\_dynamics()} plots a single simulation run, as
returned from the \texttt{run\_simulation()} function. This function is
only provided as a convenience function to provide a way to easily see
the results of a simulation run. An example plot resulting from this
function is shown in @ref(fig:plot-dynamics).

\textbackslash begin\{figure\}

\{\centering \includegraphics[width=350px]{figures/ug_three_strains_dynamics}

\}

\textbackslash caption\{Results of a simulation run shown using the
function \(plot_dynamics()\). In this case, there were three strains per
functional group, though strains within functional groups had identical
properties in this example. CB\_1 = cyanobacteria strain 1; SB\_1 =
sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria
strain 1. Here we show the figure produce by \(plot_dynamics()\), which
produces figures not with all features ideal for
publication.\}\label{fig:plot-dynamics} \textbackslash end\{figure\}

@@

\begin{quote}
\{\textgreater\textgreater This is the main section of the article and
the reviewers weight the description here appropriately Indicate in what
way new research questions can be pursued as a result of the software
(if any). Indicate in what way, and to what extent, the pursuit of
existing research questions is improved (if so). Indicate in what way
the software has changed the daily practice of its users (if so).
Indicate how widespread the use of the software is within and outside
the intended user group. Indicate in what way the software is used in
commercial settings and/or how it led to the creation of spin-off
companies (if so).\textless\textless\}
\end{quote}

The open source implementation and extension of the model used in
\protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) provides the means of
reproducing the results published while at the same time provides the
means of doing unique, innovative, and important investigations of how
ecosystems respond to environmental change, and how biodiversity may
modulate this response.

@@

Here we evidence the impact of the \emph{microxanox} package by
describing three use cases and then by describing how the package can be
a starting point for models of other ecosystems. The first two use cases
are described in detail (including the code for reproducing them) in the
\emph{User Guide} and the \emph{Partial Reproduction} vignettes. The
third is taken from \protect\hyperlink{ref-REF_NEEDED}{REF NEEDED}
(\protect\hyperlink{ref-REF_NEEDED}{2222}).

@@

The study of \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) includes simulations of the
effect of oxygen diffusivity (an environmental driver, in the sense that
it affects the ecosystem but is not affected by it) on the ecosystem
state (oxic or anoxic). The \emph{microxonax} package contains
functionality to make a specific temporal pattern of change in the
oxygen diffusivity. As well as allowing individual simulations during
which oxygen diffusivity varies, this functionality forms the basis of
the temporal method for finding stable states.

An example of this functionality is given in the \emph{Partial
Reproduction} vignette, which we briefly show here (Figure
@ref(fig:uc1). The example is composed of a single simulation, at the
beginning of which the system is in the oxic state with high abundance
of cyanobacteria. Oyxgen diffusivity is then slowly decreased and
eventually, around hour 30'000 the system switches to the anoxic state,
with high abundance of both sulfur bacteria types. The oxygen
diffusivity is then increased and at around hour 38'000 the system
abruptly switches back to the oxic state.

@@

\begin{verbatim}
#> 
#> Attaching package: 'dplyr'
#> The following objects are masked from 'package:stats':
#> 
#>     filter, lag
#> The following objects are masked from 'package:base':
#> 
#>     intersect, setdiff, setequal, union
\end{verbatim}

\textbackslash begin\{figure\}

\{\centering \includegraphics[width=350px]{/Users/owenpetchey/Desktop/microxanox/microxanox/inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching}

\}

\textbackslash caption\{The temporal dynamics of the ecosystem model
when an environmental condition (here parameter \emph{a}, the oxygen
diffusivity) changes. Plot of the stable states of the simulation runs
under different oxygen diffusivity. In this simulation there is only one
strain in each functional group. CB\_1 = cyanobacteria strain 1; SB\_1 =
sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria
strain 1. Here we show a figure adapted from the output of the
\(plot_dynamics()\) function.\}\label{fig:uc1}
\textbackslash end\{figure\}

@@

The package contains a function to extract summary features of ecosystem
responses to environmental change, such as the amount of hysteresis
displayed by the ecosystem. Hysteresis is a key feature of ecosystem
responses to environmental change, because it is related to how
difficult it can be to reserve the effects of environmental change
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). The
amount of hysteresis is measured as the extent of the environmental
condition (here oxygen diffusitivity) for which there were two stable
states. I.e. it is the extent of the environmental conditions for which
historical conditions play an important role in determining the current
system state (a definition of hysteresis).

@@

The results show that the amount of hysteresis depends greatly on the
combinations of organisms present (Figure @ref(fig:uc2)). For example,
with only the CB (cyanobacteria) present, there was no hysteresis. In
contrast, the presence of both CB and SB (sulfate reducing bacteria) led
to a large amount of hysteresis. (These results are also given in the
\emph{Partial Reproduction} vignette.)

@@

{\centering \includegraphics[width=150px]{/Users/owenpetchey/Desktop/microxanox/microxanox/inst/manuscript/figures/gen_uc2_user_guide_hysteresis}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

>>>>

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\usepackage[normalem]{ulem}
% avoid problems with \sout in headers with hyperref:
\pdfstringdefDisableCommands{\renewcommand{\sout}{}}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

%% \documentclass[preprint,12pt, a4paper]{elsarticle}

@@

\begin{document}

\begin{frontmatter}

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug%
  %
  \fnref{Corresponding Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey}
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \affiliation[University of Zürich]{Department of Evolutionary
Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Corresponding Author}
    \fntext[2]{Equal contribution}

  \begin{abstract}
  Ca. 100 words.
  \end{abstract}
    \begin{keyword}
    reproducibility; regime shift; \sep 
    reproducibility; regime shift;
  \end{keyword}

 \end{frontmatter}

@@

\begin{itemize}
\tightlist
\item
  \textbf{\textcolor{red}{update links}}
\item[$\boxtimes$]
  \sout{\textbf{\textcolor{red}{update cross references}}}
\item
  \textbf{\textcolor{red}{check keywords}}
\item[$\boxtimes$]
  \sout{\textbf{\textcolor{red}{add more references}}}
\item
  \textbf{\textcolor{red}{expand on Impact and Conclusion section}}
  \pagebreak
\end{itemize}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule()
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & CC BY 4.0 \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
& & \\
& & \\
& & \\
& & \\
& & \\
C8 & If available Link to developer documentation/manual &
\textcolor{red}{TO BE ADDED} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Rainer@krugs.de \\
\bottomrule()
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
important for developing hypotheses that can then be tested in empirical
studies. One area of ecology in which simple models have played a
influential role is the area of how ecosystems respond to gradual change
in an environmental driver \citep{Scheffer2001}. An environmental driver
is a environmental condition that affects an ecosystem, but is assumed
to not be affected by the ecosystem, such as the rate of nutrient input
into a lake.

It is conceivable that an ecosystem state, such as the total biomass of
a particular type of bacteria, may remain unchanged when an
environmental driver changes. It is also possible that the ecosystem
state changes gradually. It is also possible that the ecosystem state
changes abruptly to a new state that is difficult to recover from
\citep{Scheffer2001}. This possibility for abrupt, perhaps catastrophic
changes that are difficult to reverse causes considerable concern
\citep{Collins2021, Northrop2021, Vandermeer2019}.

An example where a gradual change of an environmental variable causes an
abrupt change of the system is the switch from an aerobic (oxygen is
available for metabolism) to anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. Three types of microbes occur in the
model: cyanobacteria (CB) dominating the oxic state, and two types of
sulfur bacteria that dominate the anoxic state (sulfate reducing
bacteria (SB) and phototrophic sulfur bacteria (PB)). This system has
been investigated by \citet{Bush2017} in a simulation study of a
mathematical model. They showed that gradual change in the rate at which
oxygen could diffuse into the ecosystem (termed the oxygen
diffusitivity) could cause catastrophic changes in the ecosystem state
that would be difficult to reverse.

One feature of the study by \citet{Bush2017} was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves open the question of if and how biodiversity
within these types (i.e.~functional groups) of bacteria, in their model,
affects the ecosystem response to environmental change. This limitation
is not specific to the study of \citet{Bush2017}, there are few if any
studies of the effects of biodiversity on abrupt transitions between
ecosystem states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change \citet{Limberger2022}, and to base our work on the
work and model of \citet{Bush2017}. It was with this goal in mind that
we developed the \emph{microxanox} package. The first stage of
development was to write code from scratch (as there was no available
code to start from) and to confirm that this new implementation would
reproduce the previously published results. The resulting reproduction
is available as one of the package vignettes:
\href{LINK\%20NEEDED}{vignette Partial reproduction of Bush et al}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on \citet{Bush2017}
(See \citet{Bush2017} for a detailed discussion of the model). At the
core of the simulations is a set of ordinary differential equations
(specified in the function \texttt{bushplus\_dynamic\_model()}, though
this function need not be directly called). There are functions for
running individual simulations and for running a set of simulations
across, for example, a range of environmental conditions.

@@

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it easy to reproduce results (see
the supplement to \citet{Limberger2022} for an example of how this is
used). As such, all the parameters needed to run a simulation or find a
stable state (i.e.~the final state of the ecosystem) are contained in a
single object (which can easily be created using included functions).
This parameter object is given to a function that runs the simulations
and returns the results. The returned results object is identical to the
parameter object but with an additional slot named \texttt{results},
which contains the results of the run. Thus the returned results object
contains the simulation conditions (parameters) as well as the results,
and can be used to run the simulation again. This promotes
reproducibility and makes incremental changes of individual parameters
and re-running the simulations straightforward.

@@

A typical simulation would look as shown in \ref{fig:runsim_example}.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package \citep{Soetaert2010}. The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains
among other things the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed please see the
\emph{User Guide} which accompanies the package or is available at
\href{@LINK_NEEDED}{User Guide}

@@

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the \citet{Bush2017} study. The second runs two
simulations, with step-wise and slowly temporally increasing oxygen
diffusivity, and the other with step-wise and slowly decreasing or
decreasing oxygen diffusivity. During this temporal environmental
change, the state of the system was recorded just before change to a new
oxygen diffusivity (we term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

The replication method is implemented in the function
\texttt{run\_replication\_ssfind()} which takes a parameter object as
returned by the function \texttt{new\_replication\_ssfind\_parameter()}
and the number of cores for multithreading the simulation. As the
multithreading uses the package function \texttt{mclapply()} from the
package \texttt{parallel} \citep{RCoreTeam2022}, the multthreading only
works on Linux and Mac. It is planned to move to \texttt{parLapply()}
\citep{RCoreTeam2022} in a future release.

@@

The function \texttt{plot\_dynamics()} plots a single simulation run, as
returned from the \texttt{run\_simulation()} function. This function is
only provided as a convenience function to provide a way to easily see
the results of a simulation run. An example plot resulting from this
function is shown in \ref{fig:plot-dynamics}.

{\centering \includegraphics[width=350px]{figures/ug_three_strains_dynamics}

\caption{Results of a simulation run shown using the function $plot_dynamics()$. In this case, there were three strains per functional group, though strains within functional groups had identical properties in this example. CB  extunderscore1 = cyanobacteria strain 1; SB extunderscore1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show the figure produce by $plot_dynamics()$, which produces figures not with all features ideal for publication.}\label{fig:plot-dynamics}
\end{figure}

@@

The open source implementation and extension of the model used in
\citet{Bush2017} provides the means of reproducing the results published
while at the same time provides the means of doing unique, innovative,
and important investigations of how ecosystems respond to environmental
change, and how biodiversity may modulate this response.

@@

Here we evidence the impact of the \emph{microxanox} package by
describing three use cases and then by describing how the package can be
a starting point for models of other ecosystems. The first two use cases
are described in detail (including the code for reproducing them) in the
\emph{User Guide} and the \emph{Partial Reproduction} vignettes. The
third is taken from \citet{REF_NEEDED}.

@@

The study of \citet{Bush2017} includes simulations of the effect of
oxygen diffusivity (an environmental driver, in the sense that it
affects the ecosystem but is not affected by it) on the ecosystem state
(oxic or anoxic). The \emph{microxonax} package contains functionality
to make a specific temporal pattern of change in the oxygen diffusivity.
As well as allowing individual simulations during which oxygen
diffusivity varies, this functionality forms the basis of the temporal
method for finding stable states.

An example of this functionality is given in the \emph{Partial
Reproduction} vignette, which we briefly show here (Figure
\ref{fig:uc1}. The example is composed of a single simulation, at the
beginning of which the system is in the oxic state with high abundance
of cyanobacteria. Oyxgen diffusivity is then slowly decreased and
eventually, around hour 30'000 the system switches to the anoxic state,
with high abundance of both sulfur bacteria types. The oxygen
diffusivity is then increased and at around hour 38'000 the system
abruptly switches back to the oxic state.

@@

\begin{verbatim}
#> 
#> Attaching package: 'dplyr'
#> The following objects are masked from 'package:stats':
#> 
#>     filter, lag
#> The following objects are masked from 'package:base':
#> 
#>     intersect, setdiff, setequal, union
#> Warning in knitr::include_graphics(fig): It is highly recommended
#> to use relative paths for images. You had absolute paths: "/
#> Users/rainer/git/0.RPackages/microxanox/inst/manuscript/figures/
#> gen_uc1_partrep_temporal_state_switching.pdf"
\end{verbatim}

{\centering \includegraphics[width=350px]{/Users/rainer/git/0.RPackages/microxanox/inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching}

\caption{The temporal dynamics of the ecosystem model when an environmental condition (here parameter *a*, the oxygen diffusivity) changes. Plot of the stable states of the simulation runs under different oxygen diffusivity. In this simulation there is only one strain in each functional group. CB   extunderscore1 = cyanobacteria strain 1; SB extunderscore1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show a figure adapted from the output of the $plot_dynamics()$  function.}\label{fig:uc1}
\end{figure}

@@

The package contains a function to extract summary features of ecosystem
responses to environmental change, such as the amount of hysteresis
displayed by the ecosystem. Hysteresis is a key feature of ecosystem
responses to environmental change, because it is related to how
difficult it can be to reserve the effects of environmental change
\citep{Scheffer2001}. The amount of hysteresis is measured as the extent
of the environmental condition (here oxygen diffusitivity) for which
there were two stable states. I.e. it is the extent of the environmental
conditions for which historical conditions play an important role in
determining the current system state (a definition of hysteresis).

@@

The results show that the amount of hysteresis depends greatly on the
combinations of organisms present (Figure \ref(figL:uc2)). For example,
with only the CB (cyanobacteria) present, there was no hysteresis. In
contrast, the presence of both CB and SB (sulfate reducing bacteria) led
to a large amount of hysteresis. (These results are also given in the
\emph{Partial Reproduction} vignette.)

\begin{verbatim}
#> Warning in knitr::include_graphics(fig): It is highly recommended
#> to use relative paths for images. You had absolute paths: "/
#> Users/rainer/git/0.RPackages/microxanox/inst/manuscript/figures/
#> gen_uc2_user_guide_hysteresis.pdf"
\end{verbatim}

@@

{\centering \includegraphics[width=150px]{/Users/rainer/git/0.RPackages/microxanox/inst/manuscript/figures/gen_uc2_user_guide_hysteresis}

@@

\renewcommand\refname{References}
\bibliography{bibliography.bib}
